# JOUR405: Statistics for Journalists
## Midterm Exam - Spring 2025

Name: Hayden Whirley

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences. Start by loading the tidyverse and any other libraries you think you might need.

```{r}
library(tidyverse)
```


## Part 1: Restaurant Health Inspections (15 points)

You want to understand how restaurants in Montgomery County are performing on health inspections. The first dataset contains restaurant health inspection scores for restaurants in Montgomery County. The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv` and complete these tasks:
```{r}
moco_restaurants <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Tasks:
1. Calculate the mean and standard deviation of compliance scores (5 points)
```{r}
moco_restaurants |> summarize(mean = mean(compliance_score), sd = sd(compliance_score))

```

2. Create a histogram of the compliance scores with a vertical line showing the mean (5 points)
```{r}
moco_restaurants |> ggplot()+
  geom_histogram(aes(x=compliance_score), binwidth = 10) +
  geom_vline(aes(xintercept = mean(compliance_score)),color="blue", size =1)
```

3. Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? (5 points).

The dataset is very skewed to the right, meaning the majority of restaurants in Montgomery County have a high to perfect health compliance score (the highest being 100). The middle 50% of the restaurants' scores fall within 5.82 standard deviations above or below the mean of 96.3, so majority have scores around 90 or higher. There are a few outliers in the 40s, 50s, 60s. I think the story could be that restaurants are doing well with health code because the data is very skewed and the average compliance score is 96, or the story could focus on those few outlier restaurants that aren't up to par with the others.

## Part 2: High School Athletics (25 points)

You are reporting a story about high school sports participation in Maryland and want to see if there are differences between boys and girls. The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and sex. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv` and complete these tasks:

```{r}
hs_participation <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
```

### Tasks:
1. Calculate the correlation between boys' and girls' participation (5 points)
```{r}
hs_participation |> summarize(correlation = cor(boys, girls, method="pearson"))
```


2. Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. (5 points)
```{r}
totals <- hs_participation |> mutate(
  total = (boys + girls), 
  girls_pct = (girls/total)*100
)
```


3. Create a scatterplot showing this relationship, adding a line of best fit (5 points)
```{r}
totals |> 
  ggplot()+
  geom_point(aes(x = total, y = girls_pct)) +
  geom_smooth(aes(x = total, y = girls_pct), method = "lm") +
  labs(title = "Total Participation vs. % Girls", subtitle = "What's the relationship?",  x = "total", y = "girls_pct")

```


4. In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

The correlation coefficient of 0.98 between girls and boys shows a very strong positive relationship. As the number of boys participating in sports increases in a county, the number of girls does as well at a near proportional rate.

On the other hand, the scatterplot shows a weak negative relationship, meaning that as the total number of students in sports increases, the percentage of girls participating tends to decrease slightly. The school districts below the line visibly have less than 50% girls in sports, and even more concerning, the two outlier school districts that are under that line but have the largest total number of students in sports should be examined further.


## Part 3: Public Transit Ridership (20 points)

You are investigating public transit ridership in the Washington, D.C. area and want to understand the patterns of daily bus and rail ridership. The third dataset contains daily bus and rail ridership totals from WMATA for the past year. Load the data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv and do the following:
```{r}
WMATA <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Tasks:
1. Calculate the average bus and rail ridership and standard deviation using summarize() (5 points)
```{r}
WMATA |> summarize(mean = mean(bus), sd = sd(bus))

WMATA |> summarize(mean = mean(rail), sd = sd(rail))
```

2. Using the process you used in class, take a random sample daily ridership numbers and calculate the sample means and deviations for bus and rail. The number in the sample is up to you, but explain why you chose what you did. Compare this to the stats you generated in step 1. (5 points)
```{r}
set.seed(1234) # Setting seed for reproduceability

sample5 <- WMATA |>
  sample_n(5)

sample10 <- WMATA |>
  sample_n(10)

```

```{r}
sample5 |> summarize(mean = mean(bus), sd = sd(bus)) sample5 |> summarize(mean = mean(rail), sd = sd(rail))

sample10 |> summarize(mean = mean(bus), sd = sd(bus)) sample10 |> summarize(mean = mean(rail), sd = sd(rail))
```


3. Using group_by() and summarize(), calculate the means for bus and rail ridership for each weekday. Describe the overall pattern of ridership for bus and rail - which days stand out and why? Are there differences between bus and rail in the standard deviation values? (10 points)

```{r}

```



## Part 4: Maryland Car Theft Rates (20 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:
```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Tasks:
1. Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate (5 points)
```{r}
car_thefts |> mutate(rate=)
```


2. Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median, and what percentage of all car thefts occur in those counties? (5 points)


3. Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? (10 points)


```{r}

```

## Part 5: Data Analysis Scenario (20 points)

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences (no code!) explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)

I would take the mean response times for each month to identify overall trends and see how much the time has actually decreased. I would also look for standard deviation to see if reponse times have become more inconsistent.

2. What visualizations would help readers understand the trends? (5 points)

I think a histogram would help visualize the distribution of reponse times to see if there has been a large skew or shift over the months of the past year.

3. What additional context or data would you need to make this a complete story? (5 points)

To make this a complete story, I would need historical response time data from previous years to compare whether the past year's increase is unusual or part of a longer trend. I think it would be useful to take the average reponse time for the whole year and compare it to years prior.


### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Before submitting, clear your environment and run the entire notebook

Good luck!
