---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Hayden Whirley"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

Based on the correlation matrix, voter turnout shows the strongest positive correlation with the percentage of the white population (0.69), and the strongest negative correlation with the unemployment rate (-0.74). I think these variables are likely predictors in this model. Some independent variables show strong correlations with eachother as well.  For example income and the percentage of residents with a bachelor's degree have a strong positive correlation (0.75). Median age and education level (0.04) as well as median age and income (-0.21) show very weak or no meaningful correlation.

Looking at the scatter plots, pct unemployed seems to have one of the strongest linear relationships with voter turnout as well as the white population. All the rest of the variables seem to have moderate to weak relationships, but it's kind of hard to tell exactly where they lie. Population and pct of people with a bachelors degree or higher seem to be the weakest linear relationships though. 


```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

I think model does a strong job of predicting voter turnout. The Multiple R-squared value is 0.8539, which means that about 85% of the variation in voter turnout is explained by the six variables in the model. The Adjusted R-squared is 0.8023, which is a bit lower but still high, so the model still has its strength even after adjusting for the number of variables. The model’s overall significance is shown by the F-statistic of 16.55 and a very small p-value (2.999e-06), so the set of variables do collectively contribute to explaining turnout variation.

The residual standard error of 0.02174 means that the model’s turnout predictions are usually off by about 2.2 percentage points, which isn’t a large error. Overall, the model does a good job, but I think it could be made even better by removing or adjusting the variables that don’t have much impact like population, percent with a bachelor’s degree or higher.

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

The full model still performs the best, with an adjusted R-squared of 0.8023, meaning it explains about 80% of the variation in voter turnout. The other models—Model 2, 3, and 4—have lower adjusted R-squared values of 0.615, 0.649, and 0.657, all pretty similar and showing they don’t explain turnout as well. I tested if using just pct_unemployed and pct_white, which had the strongest correlation to voter turnout, would work well in their own model. However, it performed worse. I then grouped all the variables with moderate to weak correlation together in model 3. While Model 4, which combined variables with different correlations, performed slightly better, the improvement was very minimal. In the end, none of the simpler models outperformed the full model, meaning the full model remains the best choice for predicting turnout. I think this shows that simplifying the model too much reduces its accuracy, making it less effective.

```{r}
# Model 2
model2 <- lm(turnout ~ pct_unemployed + pct_white, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ population + income + pct_bachelors_higher + median_age, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ median_age + pct_white + pct_bachelors_higher, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

I think the focus of an election or voter participation story based on this analysis could be the variables that most impact voter turnout in Maryland counties, such as unemployment, and the percentage of white residents. These factors were shown to have the strongest statistical relationships with voter turnout, suggesting that counties with lower unemployment rates and larger white populations tend to have higher voter participation. A story could highlight how economic and racial demographics influence civic engagement and explore why certain communities may face barriers to voting.

Some of these potential barriers to explore further could possibly be limited public transportation and infrastructure may prevent people from reaching polling places in low-income communities, especially for those who can’t afford a car or reliable transportation due to unemployment. I'd also look into whether language barriers in multilingual communities—often made up of people of color, including Hispanic and immigrant populations—limit access to clear voter information, which could discourage participation among these residents. As our analysis showed that  counties with higher percentages of white residents tend to have higher voter turnout. 
