---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The dataset has information on White House employees categorized by name, status (employee or detailee), salary, and job title. From a reporting standpoint, the data highlights differences in salary levels based on roles and employment status. I notice that some employees have a salary of 0, I'm not sure if that's because there is no data available, or if they are unpaid. This could pose problems for salary-based analysis, such as calculating averages or distributions.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(salary), median_salary = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary = mean(salary), median_salary = median(salary), min_salary = min(salary), max_salary = max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The average salary of White House employees is about $107,036, while the middle salary is $110,500. This suggests that most employees earn around this amount, but some lower salaries bring the overall average down slightly. Some employees have a recorded salary of $0, which could mean they are unpaid or their salary data is missing. These 0's are outliers because the next lowest recorded salaries are significantly higher at $20,000, $36,000, and $55,924. The highest-paid employee makes $251,258, with the job title of "Associate Counsel."

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I think an interesting way to tell the story of this data would be to explore the wide pay disparity among White House employees, where salaries range from $0 to $251,258 despite an average of $107,036. This raises questions about who holds unpaid positions, why some employees earn nothing (or very little), and whether these roles are accessible to people from diverse economic backgrounds. The next step would be to research the job titles and responsibilities tied to the lowest and highest salaries, as well as investigate whether unpaid positions impact diversity and opportunity within the White House staff.
