Jour405 Final Project Notes
Hayden Whirley 

## Final Project

***add csv file into repository

RUBRIC:

Complete submission (all files are in the proper folder) - 10 points

Notebook runs without error (or with errors explained) - 15 points

All code blocks are preceded by some narrative explaining what you're doing and why, and followed by an evaluation of the results - 20 points

Your notebook calculates both descriptive statistics and at least two additional statistical techniques that we've covered this semester - 20 points

You notebook must have at least one visualization of your choice that shows off a finding or result - 10 points

Your notebook must include a final section where you draft a story pitch, backed by your analysis. It needs to discuss your main findings, potential problems, and provide a reporting plan - 25 points

## Week 13 Assignment: Trials, Errors, and Lessons

In this analysis, I examine the relationship between the success of streaming services and the performance of movie theaters. My hypothesis is that as streaming services have become more successful in recent years, movie theaters have experienced a decline. The growing convenience of at-home viewing—offering instant access to movies and TV shows—may be contributing to a measurable decline in theater attendance and revenue.

I chose Hulu as a proxy for a leading U.S. streaming service and used a Domestic Movie Theatrical Market Summary as my source for box office data. I also included data from Box Office Mojo/IMDbPro for the number of movie releases each year, which can also play a factor in affecting the number of tickets sold and overall box office revenue. To maintain consistency and avoid the distortions caused by the COVID-19 pandemic, I am focusing on data up to 2019—when streaming services had already gained significant popularity.

***edit based on what he said
For my analysis, I will create a scatterplot to visualize the relationship between the variables, calculate the correlation coefficient, analyze the percent change across years, and perform a t-test to assess the statistical significance of the observed trends.

## Load the Tidyverse
```{r}
library(tidyverse)
```


## Loading the Data
```{r}
hulu_theaters <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTvHYClYHUErcO_da-Lfl_lTBU7oiw6OTImGqXwNDHUsPKhEoSqh7fDGOAFy8QxPoueFxkQOrZKeoKi/pub?gid=1238930269&single=true&output=csv")
```

## Create a Scatterplot for Subscribers/Tickets Sold
```{r}
hulu_theaters |>
  ggplot(aes(x = hulu_subscribers, y = mt_tickets)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Hulu Subscribers", y = "Movie Theater Tickets Sold", title = "Scatterplot of Hulu Subscribers vs Movie Theater Tickets Sold")
```


## Correlation Coefficient
```{r}
hulu_theaters |> summarize(correlation = cor(hulu_subscribers, mt_tickets, method = "pearson"))
```
s
The scatterplot shows a negative but relatively weak relationship between the number of Hulu subscribers and the number of movie theater tickets sold. The correlation coefficient indicates a moderate negative relationship, meaning that as Hulu subscriptions tend to increase, movie theater ticket sales tend to decrease. Since the correlation is not very strong, the relationship is not consistent or highly predictive, so other factors likely could influence movie theater ticket sales as well. The weak-to-moderate strength of the correlation means that Hulu subscriptions are only partially related to changes in movie ticket sales and should not be seen as the sole cause of decline in theater attendance.


## Create a Scatterplot for Revenue
```{r}
hulu_theaters |>
  ggplot(aes(x = hulu_revenue, y = mt_revenue)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Hulu Revenue", y = "Movie Theater Revenue", title = "Scatterplot of Hulu Revenue vs Movie Theater Revenue")
```


## Correlation Coefficient
```{r}
hulu_theaters |> summarize(correlation = cor(hulu_revenue, mt_revenue, method = "pearson"))

```

These results show a moderate positive correlation (0.638) between Hulu revenue and movie theater revenue. This means that as Hulu’s revenue increased, movie theater revenue also tended to increase. Rather than competing directly, this suggests the two may have grown together.


## Create a Scatterplot for Movie Releases
```{r}
hulu_theaters |>
  ggplot(aes(x = movie_releases, y = mt_tickets)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "# of Movie Releases", y = "Movie Theater Tickets Sold", title = "Scatterplot of # of Movie Releases vs Movie Theater Tickets Sold")
```


## Correlation Coefficient
```{r}
hulu_theaters |> summarize(correlation = cor(movie_releases, mt_tickets, method = "pearson"))
```
This data doesn't support that my original thought that more movie releases lead to an increase in tickets sales. The weak negative correlation (-0.26) shows the opposite—as the number of movie releases increases, ticket sales slightly decrease. While the trend isn't strong enough to make a definitive claim, it suggests that increasing the number of films in theaters does not correspond to higher overall attendance. This could mean that other factors such as changing consumer habits or economic constraints are pulling attention away from theaters altogether, regardless of how many movies are released.


## Percent Change
Here I asked AI to calculate in R, using the tidyverse, the pct change between each year in each given column. I wasn't sure about the most efficient way to do this.
```{r}
hulu_theaters <- hulu_theaters %>%
  arrange(year) %>%
  mutate(
    pct_change_hulu_subscribers = (hulu_subscribers - lag(hulu_subscribers)) / lag(hulu_subscribers) * 100,
    pct_change_hulu_revenue = (hulu_revenue - lag(hulu_revenue)) / lag(hulu_revenue) * 100,
    pct_change_mt_tickets = (mt_tickets - lag(mt_tickets)) / lag(mt_tickets) * 100,
    pct_change_mt_revenue = (mt_revenue - lag(mt_revenue)) / lag(mt_revenue) * 100,
    pct_change_movie_releases = (movie_releases - lag(movie_releases)) / lag(movie_releases) * 100
  )
```

- Hulu subscribers and hulu revenue are all positive increases each year, some smaller than others though.
- Movie theater tickets sold and revenue numbers do have some negative pct changes but no drastic decreases, just small percentages. 
- So we can't really say that movie theater success was increasing every year as much as Hulu was.


## T-Test
```{r}

```

##SUMMARY HERE

##CONCLUSION
was my hypothesis correct?


Week 12 Assignment: Tests I Could Run 

- correlation coefficient
- scatter plots
- t-test
- % change between the years

*look at past activities/homeworks

Relationship between streaming service subscriptions and movie theater attendance *in the United States
My hypothesis: increase in streaming service subscriptions → decline in movie theater attendance
How strong is the correlation?
Weak negative, weak positive, strong negative, strong positive 
Need to pick some measure of movie attendance, whether an average or some other metric
Similar work: https://filmgrail.com/blog/impact-of-streaming-on-cinema-attendance/ 

Notes from Prof. Willis for reference:
You'd need to pick some measure of movie attendance, whether an average or some other metric. See if you can find similar work.
I think that you could use AMC as a proxy for "movie theaters" in general, but not sure you can get anything below the national level (or maybe worldwide), unfortunately. As these kinds of figures are literally business secrets, they're not widely available to the public.
The last source you linked to is probably better since it's U.S. and has annual figures. Not sure I would use post-2019 numbers, though, because of COVID's impact.
Similarly, you could use Netflix as a proxy for streaming services, particularly prior to 2020, and get revenue numbers from the company's annual reports: https://ir.netflix.net/financials/annual-reports-and-proxies/default.aspx. Those might also have subscriber totals.

"Good start here, and a couple of thoughts. I wouldn't toss out Netflix numbers, but Hulu might make a better comparison (it's almost entirely US-based in terms of subscribers) to domestic movie figures. You'll need to get your data in consistent formats, including some standard amount for finances (whether that's thousands or the actual number) so you can make comparisons more easily.

Ideally your data covers the same time period, so see how much of 2010-2019 period you can get (that's really when streaming was taking off). You also might consider trying to track down the number of domestic movie releases per year in case that's a potential factor - fewer movies would logically result in less attendance.

This week's assignment is mostly about what kinds of tests you could run, and you're thinking of the right things: correlation to start, some scatterplots. Beyond that, we'll be learning a new technique next week that should provide some opportunities for you."



My Compiled Data List Spreadsheet:
https://docs.google.com/spreadsheets/d/1cQdo7W6Bd9qe-0x38SEb0gBwgAN1J1d9OYCxID_2poc/edit?usp=sharing 



Streaming Service Data

Netflix Annual Reports
https://ir.netflix.net/financials/annual-reports-and-proxies/default.aspx 
*Global, domestic available only for some years
Netflix’s yearly annual reports ranging from 2002-2024 that include revenue numbers, net income, global streaming memberships, along with much other helpful financial and membership data. They also have domestic data available as well for those categories. It’s important to note the amounts are in thousands or millions. This source also has similar netflix data compiled into tables: https://prioridata.com/data/netflix-statistics/  

Hulu Annual Reports
https://www.businessofapps.com/data/hulu-statistics/ 
*U.S. 
Business of Apps provides detailed statistics and data about Hulu, focusing on its financial performance, subscriber count, and market growth. It covers various aspects such as revenue, number of subscribers, subscriber growth, and other key metrics, with the latest data available for 2024. The time period spans from 2010 to 2024, but some metrics are only 2019-2024. Most importantly the data I would want (subscribers & revenues) range from 2010/2011 to 2024.

Notes:

I was considering getting the same metrics from HBO Max to get a wider picture, but then realized they didn’t launch until 2020, so the post-covid data would not be helpful.

It seems the Netflix data is global, but only some years have domestic data available for revenue & subscription totals, or it's combined with Canada. I’m having trouble determining if Hulu’s data is domestic or international. After googling, Hulu is a U.S. company, but your confirmation would be helpful. Hulu then might be better suited to compare with the movie theater data I was able to find, which is mostly domestic.


Movie Theater Data

Domestic Theatrical Market Summary 1995 to 2025
https://www.the-numbers.com/market/ 
*U.S.
This source features a comprehensive table of U.S. movie ticket sales and box office data from 1995 to 2025. It covers metrics like annual ticket sales, total gross revenue, and average ticket prices. The data is in table format on the site and can be manually entered into Excel or CSV. It’s a valuable resource to use for analyzing movie theater success or decline in the U.S. I think annual ticket sales serve as a way to measure attendance, since direct visitor data isn’t provided. It seems that finding specific data on the number of people physically entering movie theaters is unavailable, so ticket sales may be the best proxy.

Domestic Yearly Box Office 
https://www.boxofficemojo.com/year/?grossesOption=totalGrosses
*U.S.
I think this is also a good source from IMDbPro for looking at domestic movie theater revenue trends, but it doesn’t include ticket sales or attendance figures. It tracks the total annual gross revenue of films released in U.S. theaters from 1977 to 2025. The data includes metrics such as total gross, number of releases, average gross per release, and the highest-grossing film each year.

Notes:
These have been the top sources I have been able to find, I think I am leaning towards comparing the Hulu totals with Domestic Movie Theater totals from The Numbers.

Questions:
When calculating correlation and creating charts to look at the relationship between streaming service and movie theater increases/declines, I know I want to find correlation, but what other data or metrics can be calculated to determine the relationship between the two?

