Jour405 Final Project
Hayden Whirley 

## Introduction 

In this analysis, I examine the relationship between the success of streaming services and the performance of movie theaters. My hypothesis is that as streaming services have become more successful, movie theaters have experienced a decline. The growing convenience of at-home viewing—offering instant access to movies and TV shows—may be contributing to a measurable decline in theater attendance and revenue.

I chose Hulu annual reports as a proxy for a U.S. streaming service and used a Domestic Movie Theatrical Market Summary as my source for box office data. I compiled a dataset that includes Hulu subscribers each year, domestic movie theater tickets sold each year, and revenues for both, as variables to measure the success of both these entertainment platforms. I also included data from Box Office Mojo/IMDbPro for the number of movie releases each year, which can also play a factor in affecting the number of tickets sold and overall box office revenue. 

To maintain consistency and avoid the distortions caused by the COVID-19 pandemic, I am focusing on data up to 2019—when streaming services had already gained significant popularity. Movie theaters have been around since the late 19th century, while Hulu launched in 2008 and has become a leading streaming service in the United States. The purpose of this analysis is to determine if streaming service usage has influenced movie-theater going behavior in recent years as technology and convenience have become more prevalent.

For my analysis, I will calculate the correlation coefficient, create a scatterplot to visualize the relationship between the variables, analyze the percent change across years, perform a t-test to assess the statistical significance of the changes in average movie ticket sales, and _____ a multiple regression model to see if my variables can predict eachother.


## Load the Tidyverse
```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Load My Data
```{r}
hulu_theaters <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTvHYClYHUErcO_da-Lfl_lTBU7oiw6OTImGqXwNDHUsPKhEoSqh7fDGOAFy8QxPoueFxkQOrZKeoKi/pub?gid=1238930269&single=true&output=csv")
```

First, I'm calculating the correlation coefficient to measure the strength and direction of the linear relationship between Hulu subscriptions and movie theater tickets sold each year. This helps me determine whether there's a meaningful connection between the two variables.

## Section 1 - Correlation Coefficient
```{r}
hulu_theaters |> summarize(correlation = cor(hulu_subscribers, mt_tickets, method = "pearson"))
```

This correlation coefficient (-0.54) indicates a moderate negative relationship, meaning that as Hulu subscriptions tend to increase, movie theater ticket sales tend to decrease. Since the correlation is not very strong, the relationship is not consistent or highly predictive, so other factors likely could influence movie theater ticket sales as well. The moderate strength of the correlation means that Hulu subscriptions are only partially related to changes in movie ticket sales and should not be seen as the sole cause of decline in theater attendance. A coefficient closer to -0.7 to -1 would determine a stronger correlation and we could say that these two variables have a more consistent and predictable linear relationship.



Next, I will create a scatterplot for those same variables to better visualize the relationship between Hulu subscriptions and movie theater ticket sales. This visual representation will help reveal any patterns, trends, or outliers that may not be as clear from the correlation coefficient alone.

## Section 2 - Scatterplot
```{r}
hulu_theaters |>
  ggplot(aes(x = hulu_subscribers, y = mt_tickets)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Hulu Subscribers", y = "Movie Theater Tickets Sold", title = "Scatterplot of Hulu Subscribers vs Movie Theater Tickets Sold")
```

This scatterplot shows that while there is a slight downward trend—represented by the red line of best fit indicating a negative relationship, the data points are not tightly clustered along the line. This suggests that the correlation is not very strong. The variability in the data points means that while an increase in Hulu subscriptions may be associated with a decrease in movie ticket sales, this trend isn't consistent or reliable enough to suggest a strong or direct relationship. Other external factors likely influence ticket sales, and Hulu subscription numbers alone cannot fully predict them accurately.



Here I am calculating the percent change for both Hulu subscriptions and movie theater ticket sales between each year from 2011 to 2019. I'm doing this to observe the year-to-year increases or decreases, identify any significant jumps or drops, and better understand how the trends in both variables compare over time. This helps reveal whether any changes in Hulu growth align with shifts in movie theater attendance. This serves as an additional way to highlight specific points of change that could indicate meaningful patterns or turning points.

## Section 3 - PCT Change 
*AI Use

```{r}
hulu_theaters <- hulu_theaters %>%
  arrange(year) %>%
  mutate(
    pct_change_hulu_subscribers = (hulu_subscribers - lag(hulu_subscribers)) / lag(hulu_subscribers) * 100,
    pct_change_mt_tickets = (mt_tickets - lag(mt_tickets)) / lag(mt_tickets) * 100,
  )
```

With the addition of these new pct chaneg columns, we can see that while Hulu subscribers increased every year from 2011 to 2019—sometimes significantly, such as the 100% increases in both 2012 and 2013—the number of movie theater tickets sold showed less consistency. Several years saw declines in ticket sales, including noticeable drops of around 6% in both 2014 and 2019. Although there were occasional increases, like the 7.8% rise in 2012 and 7% in 2018, overall the trend for movie ticket sales was more unstable. In summary, I can say that this contrast suggests Hulu experienced steady growth and no decreasing numbers at all, and movie theater attendance did not follow a fully positive or reliable trend. These fluctuations may point to the growing popularity of streaming services influencing consumer behavior, but we can't assume causation.



We know that Hulu subscribers increased every year, but I wanted to confirm whether movie ticket sales changed over the same period. I want to conduct a one-sample t-test to determine whether the average number of movie tickets sold during the later years (2015–2019) was significantly lower than the average during the early years (2011–2014).

**Hypotheses:**
Null: There is no decrease in movie ticket sales in the later years compared to earlier years
Alternative: Movie ticket sales have decreased in the later years

## Section 4 - T-Test 
*AI Use

```{r}
# Filter out early and later periods
early_tickets <- hulu_theaters %>%
  filter(year <= 2014) %>%
  pull(mt_tickets)

later_tickets <- hulu_theaters %>%
  filter(year > 2014) %>%
  pull(mt_tickets)

# Set the significance level for your test
alpha_level <- 0.05 # Replace with the appropriate value

# Perform a one-sample t-test
# Since we want to know if ticket sales declined (decreased), we use a one-sided test (alternative = "less")
t_test_result <- t.test(
  later_tickets,
  mu = mean(early_tickets),
  alternative = "less"
)

# Display the results
t_test_result
```

The mean number of tickets sold in the early period was 1,315,562,466, while the mean in the later period was 1,280,447,985. So the mean was lower in the second half of the data by 35,114,481 tickets, which I would say is a notable difference.

However, the results of the t-test had a t statistic of -1.523, meaning the average number of movie tickets sold in the later years is about 1.523 standard errors below the average in the earlier years, showing a moderate decrease relative to the variability in ticket sales.

The p-value is 0.1012. Since the p-value is greater than the significance level of 0.05, about double the amount, we fail to reject the null hypothesis. This means that even though the average ticket sales were lower in the later period, there is not enough statistical evidence to conclude that movie ticket sales significantly declined compared to the earlier years.




## Section 5 - Linear Regression Models


# First: Can hulu subscribers predict movie theater tickets sold?

```{r}
# Fit a simple linear regression model
model <- lm(mt_tickets ~ hulu_subscribers, data = hulu_theaters)

# View the model summary
summary(model)
```
P-value: 0.1308
Adjusted R-squared value: 0.1941
Coefficients: y-intercept: 1,332,661,684 slope for Hulu Subscribers: -3.328

The y-intercept (1,332,661,684) represents the predicted number of movie theater tickets sold when there are zero Hulu subscribers. The slope (-3.328) means that for every additional Hulu subscriber, the model predicts about 3.33 fewer movie tickets sold. This suggests a negative relationship, for example, more Hulu subscribers might be associated with fewer movie tickets sold. 

The adjusted R-squared of 0.1941 means that only 19.41% of the variation in movie ticket sales can be explained by the number of Hulu subscribers. So about 80.59% of the variation is due to other factors, not captured by this model. So, this is a weak model.

Since 0.1308 > 0.05, we fail to reject the null hypothesis. This means the negative relationship is not statistically significant — we do not have strong enough evidence to conclude that Hulu subscribers meaningfully predict movie ticket sales.


# Second: Can Hulu revenue predict movie theater revenue?

```{r}
# Fit a simple linear regression model
model <- lm(mt_revenue ~ hulu_revenue, data = hulu_theaters)

# View the model summary
summary(model)
```

P-value: 0.06452
Adjusted R-squared value: 0.3222
Coefficients: y-intercept: 10,473,582,009 slope for Hulu Revenue: 0.2492

The y-intercept ($10,473,582,009) is the baseline predicted movie theater revenue when Hulu revenue is $0. The slope (0.2492) means that for every additional $1 in Hulu revenue, the model predicts an increase of about $0.25 in movie theater revenue. So there's a positive relationship — higher Hulu revenue is associated with higher movie theater revenue in this model. 

The adjusted R-squared of 0.3222 means that about 32.22% of the variation in movie theater revenue is explained by Hulu revenue. While not super strong, it can be considered moderate.

This p-value of 0.06452 is just above the common 0.05 threshold. That means the result is not quite statistically significant, but it's close.



# Third: Can movie releases predict movie theater tickets sold?

```{r}
# Fit a simple linear regression model
model <- lm(mt_tickets ~ movie_releases, data = hulu_theaters)

# View the model summary
summary(model)
```

P-value: 0.4993
Adjusted R-squared value: -0.06561
Coefficients: y-intercept: 1,427,206,055 slope for Movie Releases: -181,204

The y-intercept (1,427,206,055) represents the estimated number of tickets sold when there are 0 movie releases. The slope (-181,204) means that for each additional movie released, the model predicts a decrease of about 181,204 movie tickets sold. That’s a negative relationship, which is odd.

The adjusted R-squared here of -0.06561 means this model doesn’t explain any meaningful variation in the data and is not useful for prediction.

Lastly, the p-value of 0.4993 is very high, which means there's no statistically significant evidence that the number of movie releases has any predictive power for ticket sales.



# Fourth: Can movie releases predict Hulu subscribers?

```{r}
# Fit a simple linear regression model
model <- lm(hulu_subscribers ~ movie_releases, data = hulu_theaters)

# View the model summary
summary(model)
```

P-value: 0.001192
Adjusted R-squared value: 0.7682
Coefficients: y-intercept: -62,492,588 slope for Movie Releases: -101,540

The y-intercept (-62,492,588) is the predicted number of Hulu subscribers when there are zero movie releases, obviously not realistic, but it's a baseline. The slope (-101,540) means that for each additional movie released, the model predicts about 101,540 fewer Hulu subscribers. This suggests negative relationship, more movies in theaters might correlate with fewer people subscribing to Hulu.

The adjusted R-squared value of 0.7682 is pretty strong, meaning t 76.82% of the variation in Hulu subscribers is explained by the number of movie releases.

The p-value of 0.001192 is much smaller than 0.05, so we can conclude that there is a statistically significant relationship between movie releases and Hulu subscribers, and this could be worth exploring further. The model predicts that more movie releases are associated with fewer Hulu subscribers.




## Story Pitch
(final section)
draft a story pitch, backed by your analysis. It needs to discuss your main findings, potential problems, and provide a reporting plan

Main Findings:
- correlation coefficient showed us that
- scatterplot let us see 
- pct change in subscribers and movie theater tickets told us 
- one sample t-test resulted in
- multiple regression model 

focus this on how i orginally thought there would be more statically signidcinat realtionship, so my conclusion will reject/counter that


Reporting Plan:

Story pitch - 

Headline -
Lede -

Who I would interview - 

Potential problems - 
